---
title: "Ground Combat Systems (GCS) Analysis"
format: html
editor: visual
---

# Objective

## Programs of Interest

Illuminate what the potential client Department spent on the following major Ground Combat Systems (GCS) programs:

-   Abrams tank (e.g., M1A1)
-   Bradley fighting vehicle (e.g., M2A1)
-   Stryker armored personnel carrier (e.g., M1130)

## Analysis Requirements

Produce an analysis of FY16 through FY20 federal contract data for these programs:

-   Summarize prominent vendors, funding/contracting offices, and agencies
-   Outline key technologies, products, and services acquired
-   Increase understanding of what future acquisitions for each of these programs may look like given:
    -   Where they are in their life cycles
    -   What potential risks may be associated with the management of the programs

## Analysis Structure

1)  Solutions to the following pain points the potential client has previously encountered in their data, including:

-   Vendor name normalization
-   Misidentified or unidentified GCS contracts and vendors

2)  Assessment of the following:

-   Spending trends
-   Key technologies
-   Programs
-   Vendors
-   Any other elements interesting about the market (e.g., current acquisition life cycle status of the GCS programs and what it means for future investments; or past/future federal budget alignment within each program).

3)  Assessment should include:

-   Conclusion with key takeaways
-   Conclusion with 1-2 interesting insights that potential client can use to inform their acquisition decisions

4)  Commented-code or a short summary of methodology used to clean/normalize, organize, and/or transform the data.

# Initialize Project

## Load packages & functions

```{r}
# packages for data manipulation
library(dplyr)
library(openxlsx2)
library(purrr)
library(stringr)
library(stringdist)
library(tidyr)
library(ggplot2)
library(scales)
library(readxl)

# packages required to access usaspending api
library(httr2)
library(jsonlite)

# source anonymous functions
source(paste0(getwd(), "/functions/clean.R"))
```

## Set parameters

```{r}
#dir_path <-paste0("/Users/", Sys.info()["user"], "/R/DATA/")
dir_path <- paste0("C:\\Users\\lmurray\\Software\\Git Local\\AnalyticsExercises")
file_name <- "ground_vehicles.xlsx"
data_location <- file.path(dir_path, file_name)
```

## Read data

```{r}
ground_vehicles <- read_xlsx(data_location)
View(ground_vehicles) # open dataframe when this chunk is run
```

# Clean data

## Identify cons of interest

```{r}
# filter dataframe based on ground combat systems program portfolio
## use toupper to normalize data to be ALL CAPS
keywords <- toupper(c("Abrams tank", "M1-A1", "M-1 Abrams", "Abrams", "M1 Abrams",
              "Bradley fighting vehicle", "Bradley", "BVF", "M2-A1", "M2A1", "M2",
              "Stryker armored personnel carrier", "M1130", "M1255 CVV", "M1130 Commander Vehicle", "M1130 CV"))
abrams <- keywords[1:5]
bradley <- keywords[6:11]
stryker <- keywords[12:17]

# create gcs_item column for ease of analysis later on
gcs_portfolio <- ground_vehicles %>% 
  filter(str_detect(toupper(Title), pattern = paste(keywords, collapse = '|'))) %>% 
  mutate(
     gcs_item = case_when(
      grepl(paste(abrams, collapse = '|'), toupper(Title)) ~ "abrams",
      grepl(paste(bradley, collapse = '|'), toupper(Title)) ~ "bradley",
      grepl(paste(stryker, collapse = '|'), toupper(Title)) ~ "stryker",
      TRUE ~ NA_character_  # Default case if none of the above conditions are met
    )
  )
```

# Solution 1: stringR/regex for normalization

To normalize data fields such as vendor name, it is often easily done with regular expressions and `stringR` functions. To identify the extent of the problem, we can first examine the number of unique `Vendor Name`s available in the data set.

## Identify problem

```{r}
unique_vendor_count <- ground_vehicles %>% 
  pull(12) %>% 
  unique()

print(length(unique_vendor_count))
```

## Apply stringR/regex solution

```{r}
ground_vehicles_cleaned <- ground_vehicles %>% 
  rename(vendor_name = 12) %>% 
  mutate(
    vendor_name_clean = toupper(vendor_name),  # convert to all caps
    vendor_name_clean = str_replace_all(vendor_name_clean, "[.,]", ""),  # remove punctuation
    vendor_name_clean = str_replace_all(vendor_name_clean, "\\bCOMPANY\\b", "CO"),  # normalize "COMPANY" to "CO"
    vendor_name_clean = str_replace_all(vendor_name_clean, "\\bCORPORATION\\b", "CORP")  # normalize "CORPORATION" to "CORP"
  )
```

## Evaluate solution

```{r}
unique_vendor_count_cleaned <- ground_vehicles_cleaned %>% 
  pull(15) %>% 
  unique()

print(paste0("There are now ", length(unique_vendor_count_cleaned), " unique vendor names in the data set, when formerly there were ", length(unique_vendor_count), "! This is a difference of ", length(unique_vendor_count)-length(unique_vendor_count_cleaned), " overlapping names, prior to normalization. Nice!"))
```

# Solution 2: query USASpending for additional data

This solution is the initial step to provide misidentified or unidentified GCS contract modifications. With this same method, other API calls to the USASpending database can be queried to find additional provide misidentified or unidentified GCS contracts and vendors.

The USASpending API (Application Programming Interface) allows the public to access comprehensive U.S. government spending data ([read more](https://api.usaspending.gov/)). This example shows how to access this API utilizing the `httr2` package.

There are limits to the external server's ability to handle requests, thus a final solution would be implementing a full data pipeline to leverage all relevant USASpending information. This can be done through several ways, including:

-   Creating an external storage and server host of a copy of the database for faster calls
-   Customizing API calls with more efficient querying

## Identify Contract/DOs

```{r}
unique_ids <- gcs_portfolio %>% 
  select("contractnumber" = 2, "deliveryorder" = 3) %>% 
  mutate(across(everything(), ~ remove_non_alphanumeric(.x))) %>% 
  unique()
raw_ids <- raw_id(unique_ids) %>% 
  unlist %>% 
  unique()
```

## Convert list to API call format

```{r}
# use ananymous function 
combined_list <- format_contract_numbers(unique_ids)
combined_list <- lapply(combined_list, na.omit)
```

## Perform API call

```{r}
# initialize list
final_con_list <- list()
```

```{r}
final_con_list$con <- transactions_request(combined_list$contract_num_formatted)
```

```{r}
final_con_list$do  <- transactions_request(combined_list$contract_num_do_formatted)
```

## Cleaned USASpending query

```{r}
usa_spending_df <- extract_results(final_con_list) %>% 
  mutate(
    award_id = str_extract(id, paste(raw_ids, collapse = "|"))
  )
View(usa_spending_df)
```

# Assessments

## Create theme

```{r}
program_colors <- c(
  abrams  = "#4B5320",  # Army Green
  bradley = "#1C1C1C",  # Army Black
  stryker = "#FFCC01"   # Army Gold
)

agency_colors <- c(
  "#4B5320",
  "#FFCC01",
  "#A7A8AA",
  "#556B2F",
  "#6B8E23",
  "#D9D9D6"
)

program_colors_high_contrast <- c(
  abrams  = "#4B5320",  # Army Green
  bradley = "#C69214",  # Army Gold (darker)
  stryker = "#5B6770"   # Steel Gray
)

theme_army <- function() {
  theme_minimal(base_size = 18) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_line(color = "grey85"),
      legend.position = "bottom",
      legend.title = element_blank(),
      plot.title = element_text(face = "bold", size = 20),
      axis.title = element_text(face = "bold"),
      plot.title.position = "plot"
    )
}

gcs_plots <- list()
```

## Spending trends

-   Award amounts in this dataset max at 2019
-   Abrams by far spent the most

```{r}
spending <- gcs_portfolio %>%
  select(fiscal_year = 5, award_amount = 7, gcs_item = 15) %>%
  mutate(gcs_item = str_to_title(gcs_item)) %>%
  group_by(fiscal_year, gcs_item) %>%
  summarise(award_amount = sum(replace_na(award_amount, 0)), .groups = "drop") %>%
  complete(fiscal_year, gcs_item, fill = list(award_amount = 0)) %>%
  mutate(
    label_y = case_when(
      gcs_item == "Abrams" ~ award_amount + 150000000,
      gcs_item == "Bradley" ~ award_amount + 80000000,
      gcs_item == "Stryker" ~ award_amount - 50000000
    )
  )

gcs_plots$spending_trends <- ggplot(
  spending,
  aes(fiscal_year, award_amount, color = gcs_item, group = gcs_item)
) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_label(
  aes(
    y = label_y,
    label = scales::label_dollar(
      scale_cut = scales::cut_short_scale()
    )(award_amount)
  ),
  size = 2.8,
  label.size = 0.15,
  fill = "white",
  color = "black",
  show.legend = FALSE
) +
  scale_color_manual(
    values = c(
      Abrams = "#4B5320",
      Bradley = "#1C1C1C",
      Stryker = "#FFCC01"
    )
  ) +
  scale_y_continuous(
    labels = scales::label_dollar(scale_cut = scales::cut_short_scale()),
    expand = expansion(mult = c(0.05, 0.25))
  ) +
  scale_x_continuous(
    breaks = sort(unique(spending$fiscal_year))
  ) +
  labs(
    title = "Annual Contract Spending by Ground Combat System",
    subtitle = "FY16-FY20 award obligations by program",
    x = "Fiscal Year",
    y = "Award Amount",
    color = "Program"
  ) +
  theme_army()
```

## Key technologies

Key take-aways:

-   Product and Service Code (PSC) is a four-digit code used by all federal government contracting activities for identifying and classifying the services and Supplies & Equipment (S&E) that are purchased under contract
-   Top 10 PSC by GCS program shows two top PSC shared by the Abrams and Bradley programs
-   More granular detail can be discern if additional time were utilized to scrape PSC definitions and identify key technologies or PSC of interest

```{r}
# select and count PSCs by GCS program
technologies <- gcs_portfolio %>% 
  select(1, "fiscal_year" = 5, "psc" = 14, 15) %>% 
  mutate(psc_cleaned = str_extract(psc, "\\(([A-Za-z0-9]{4})\\)")) %>%
  filter(!is.na(psc_cleaned)) %>%
  count(gcs_item, psc_cleaned, sort = TRUE) %>%
  group_by(gcs_item) %>%
  slice_max(order_by = n, n = 10, with_ties = FALSE) %>%
  ungroup()

# top 10 PSC by GCS program
gcs_plots$psc_code <- ggplot(technologies, aes(x = reorder(psc_cleaned, n), y = n, fill = gcs_item)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~gcs_item, scales = "free_y") +
  coord_flip() +
  scale_fill_manual(values = program_colors) +
  labs(
    x = "Product & Service Code (PSC)",
    y = "Contract Count",
    title = "Top 10 Product & Service Codes by GCS Program"
  ) +
  theme_army()
```

## Programs

-   The Department of Defense by far had the most data points available, which is representative of contract modifications or contract share

```{r}
# create agency share table
program_summary <- gcs_portfolio %>%
  select(
    gcs_item = 15,
    contracting_agency = 9
  ) %>%
  filter(!is.na(contracting_agency)) %>%
  group_by(gcs_item) %>%
  summarise(
    total_contracts = n(),
    unique_agencies = n_distinct(contracting_agency),
    primary_agency = names(sort(table(contracting_agency), decreasing = TRUE))[1],
    primary_agency_share = max(table(contracting_agency)) / n(),
    .groups = "drop"
  )

gcs_plots$agency_share <- ggplot(
  program_summary,
  aes(
    x = reorder(gcs_item, primary_agency_share),
    y = primary_agency_share,
    fill = gcs_item
  )
) +
  geom_col(show.legend = FALSE) +
  geom_text(
    aes(label = scales::percent(primary_agency_share, accuracy = 1)),
    hjust = -0.1,
    size = 5
  ) +
  scale_fill_manual(values = program_colors) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1.1)
  ) +
  coord_flip() +
  labs(
    title = "GCS Program Contracting Concentration",
    subtitle = "Share of contracts awarded by largest contracting agency (DoD)",
    x = NULL,
    y = "Share of Contract Actions"
  ) +
  theme_army()
```

## Vendors

-   L3 Technologies, Inc. had the highest award amount in 2016, but thereafter General Dynamics Land Systems, Inc. dominated in all subsequent years available in the dataset

```{r}
# summarize vendor awards
vendor_summary <- gcs_portfolio %>% 
  select(
    award_amount = 7,
    vendor = 12,
    gcs_item = 15
  ) %>%
  mutate(
    award_amount = replace_na(award_amount, 0)
  ) %>%
  group_by(vendor, gcs_item) %>%
  summarise(
    total_awards = sum(award_amount),
    .groups = "drop"
  ) %>%
  group_by(vendor) %>%
  mutate(
    vendor_total = sum(total_awards)
  ) %>%
  ungroup() %>%
  slice_max(vendor_total, n = 10) 

# calculate vendor totals for labels
vendor_labels <- vendor_summary %>%
  group_by(vendor) %>%
  summarise(
    vendor_total = sum(total_awards),
    .groups = "drop"
  )

gcs_plots$vendors <- ggplot(
  vendor_summary,
  aes(
    x = reorder(vendor, vendor_total),
    y = total_awards,
    fill = gcs_item
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    data = vendor_labels,
    aes(
      x = vendor,
      y = vendor_total,
      label = scales::label_dollar(
        scale_cut = scales::cut_short_scale()
      )(vendor_total)
    ),
    inherit.aes = FALSE,
    hjust = -0.1,
    size = 4
  ) +
  coord_flip() +
  scale_fill_manual(
    values = program_colors_high_contrast,
    labels = c(
      abrams = "Abrams",
      bradley = "Bradley",
      stryker = "Stryker"
    )
  ) +
  scale_y_continuous(
    labels = scales::label_dollar(
      scale_cut = scales::cut_short_scale()
    ),
    expand = expansion(mult = c(0, 0.2))
  ) +
  labs(
    title = "Top Vendors by Ground Combat System Award Amount (FY16-FY20)",
    subtitle = "General Dynamics Land Systems dominates the GCS vendor base",
    x = NULL,
    y = "Award Amount",
    fill = "Program"
  ) +
  theme_army()
```

# Print

```{r}
fig_path <- "figures"

#dir.create(fig_path, showWarnings = FALSE)

purrr::iwalk(
  gcs_plots,
  ~ ggsave(
      filename = file.path(fig_path, paste0(.y, ".png")),
      plot = .x,
      width = 8,
      height = 6,
      dpi = 300
    )
)
```