---
title: "Ground Combat Systems (GCS) Analysis"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

# Objective

## Programs of Interest

Illustrate funding for the following major U.S. Army Ground Combat Systems (GCS) programs contained in the portfolio data set:

-   Abrams tank (e.g., M1A1)
-   Bradley fighting vehicle (e.g., M2A1)
-   Stryker armored personnel carrier (e.g., M1130)

## Analysis Requirements

Produce an analysis of FY16 through FY20 federal contract data for the programs of interest:

-   Supplement with additional contract data
-   Summarize spending, key technologies, contracting offices, and vendors

# Initialize Project

-   Load and document required packages and named functions (`clean.R`)
-   Set parameter values (e.g., folder paths)
-   Read in the GCS portfolio Microsoft Excel file.

## Load packages & functions

```{r}
# packages for data manipulation
library(dplyr)
library(purrr)
library(stringr)
library(stringdist)
library(readxl)
library(tidyr)

# packages for data visualization
library(ggplot2)
library(scales)

# packages required to access usaspending api
library(httr2)
library(jsonlite)

# source named functions
source(paste0(getwd(), "/functions/clean.R"))
```

## Set parameters

```{r}
dir_path <-paste0("/Users/", Sys.info()["user"], "/R/DATA/")
#dir_path <- paste0("C:\\Users\\", Sys.info()["user"], "\\Software\\Git Local\\AnalyticsExercises")
file_name <- "ground_vehicles.xlsx"
data_location <- file.path(dir_path, file_name)
```

## Read data

```{r}
ground_vehicles <- read_xlsx(data_location)
ground_vehicles
```

# Clean data

Clean portfolio data set to sole Contracts, Delivery Orders, and Modifications of interest to Ground Combat Systems:

-   Filter data based on GCS portfolio and apply tags
-   Identify and normalize common string inconsistencies
-   Perform an ad hoc validation of applied changes

## Filter data & apply tags

```{r}
# filter dataframe based on ground combat systems program portfolio
keywords <- toupper(c("Abrams tank", "M1-A1", "M-1 Abrams", "Abrams", "M1 Abrams",
              "Bradley fighting vehicle", "Bradley", "BVF", "M2-A1", "M2A1", "M2",
              "Stryker armored personnel carrier", "M1130", "M1255 CVV", "M1130 Commander Vehicle", "M1130 CV"))
abrams <- keywords[1:5]
bradley <- keywords[6:11]
stryker <- keywords[12:17]

# create gcs_item column as metadata tag for each vehicle for ease of analysis later on
gcs_portfolio <- ground_vehicles %>% 
  filter(str_detect(toupper(Title), pattern = paste(keywords, collapse = '|'))) %>% 
  mutate(
     gcs_item = case_when(
      grepl(paste(abrams, collapse = '|'), toupper(Title)) ~ "abrams",
      grepl(paste(bradley, collapse = '|'), toupper(Title)) ~ "bradley",
      grepl(paste(stryker, collapse = '|'), toupper(Title)) ~ "stryker",
      TRUE ~ NA_character_
    )
  )
```

## Identify string inconsistencies

```{r}
unique_vendor_count <- ground_vehicles %>% 
  pull(12) %>% 
  unique()

print(paste0("Unique vendor names: ", length(unique_vendor_count)))
```

## Normalize strings

```{r}
ground_vehicles_cleaned <- ground_vehicles %>%
  rename(vendor_name = 12) %>%
  mutate(
    vendor_name_clean = vendor_name %>%
      str_to_upper() %>%
      str_squish() %>% 
      str_replace_all("[.,]", "") %>%
      str_replace_all("\\bCO\\b", "COMPANY") %>%
      str_replace_all("\\bCORP\\b", "CORPORATION")
  )
```

## Validate change

```{r}
unique_vendor_count_cleaned <- ground_vehicles_cleaned %>% 
  pull(15) %>% 
  unique()

print(paste0("Unique vendor names: ", length(unique_vendor_count_cleaned)))
print(paste0("Delta: ", length(unique_vendor_count)-length(unique_vendor_count_cleaned)))
```

# Query USASpending API

USASpending API (Application Programming Interface) allows the public to access comprehensive U.S. government spending data ([read more](https://api.usaspending.gov/)):

-   Access API utilizing the `httr2` package
-   Apply named functions for more efficient querying

## Identify Contract/DOs

```{r}
unique_ids <- gcs_portfolio %>% 
  select("contractnumber" = 2, "deliveryorder" = 3) %>% 
  mutate(across(everything(), ~ remove_non_alphanumeric(.x))) %>% 
  unique()

raw_ids <- raw_id(unique_ids) %>% 
  unlist %>% 
  unique()
```

## Convert list to API call format

```{r}
combined_list <- format_contract_numbers(unique_ids)
combined_list <- lapply(combined_list, na.omit)
```

## Perform API call

```{r}
final_con_list <- list()
```

```{r}
final_con_list$con <- transactions_request(combined_list$contract_num_formatted)
```

```{r}
final_con_list$do  <- transactions_request(combined_list$contract_num_do_formatted)
```

## Clean query

```{r}
usa_spending_df <- extract_results(final_con_list) %>% 
  mutate(
    award_id = str_extract(id, paste(raw_ids, collapse = "|"))
  )
usa_spending_df
```

# Perform Assessments

Analyze data by creating visuals to represent summary statistics:

-   Create consistent theme for readability and usability
-   Pivot and visualize data utilizing `dplyr` and `ggplot2`
-   Document key insights

## Create theme

```{r}
program_colors <- c(
  abrams  = "#4B5320",  # Army Green
  bradley = "#1C1C1C",  # Army Black
  stryker = "#FFCC01"   # Army Gold
)

program_colors_upper = c(
      Abrams  = "#4B5320",
      Bradley = "#1C1C1C",
      Stryker = "#FFCC01",
      Total   = "#6E6F72"
    )

agency_colors <- c(
  "#4B5320",
  "#FFCC01",
  "#A7A8AA",
  "#556B2F",
  "#6B8E23",
  "#D9D9D6"
)

program_colors_high_contrast <- c(
  abrams  = "#4B5320",  # Army Green
  bradley = "#C69214",  # Army Gold (darker)
  stryker = "#5B6770"   # Steel Gray
)

theme_army <- function() {
  theme_minimal(base_size = 18) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_line(color = "grey85"),
      legend.position = "bottom",
      legend.title = element_blank(),
      plot.title = element_text(face = "bold", size = 20),
      axis.title = element_text(face = "bold"),
      plot.title.position = "plot"
    )
}

gcs_plots <- list()
```

## Combined query

Compare portfolio data to USASpending API query:

-   Per number of observations in the `gcs_portfolio` (1,154) and `usa_spending_df` (1,697) – there are visbly relatively mhere are more mods available for these Contract numbers in USASpending
-   There are two outlier contract numbers with significantly more portfolio modifications
-   More common that one data source had the larger share of mods as compared to the other with few points directly on the dotted line where n=1

```{r}
comparison_summary <- gcs_portfolio %>%
  transmute(
    contract_number = `Award Number (PIID)`,
    modification_number = `Modification Number`,
    gcs_item
  ) %>%
  filter(!is.na(contract_number)) %>%
  distinct(contract_number, modification_number, gcs_item) %>%
  count(contract_number, gcs_item, name = "portfolio_modifications") %>%
  left_join(
    usa_spending_df %>%
      distinct(award_id, modification_number) %>%
      count(award_id, name = "usaspending_modifications"),
    by = c("contract_number" = "award_id")
  ) %>%
  mutate(
    usaspending_modifications = coalesce(usaspending_modifications, 0L),
    modification_difference =
      usaspending_modifications - portfolio_modifications
  ) %>%
  arrange(desc(abs(modification_difference)))

gcs_plots$comparison_summary <-
  ggplot(
    comparison_summary,
    aes(
      x = portfolio_modifications,
      y = usaspending_modifications,
      color = gcs_item
    )
  ) +
  geom_text(
    data = subset(comparison_summary, abs(modification_difference) > 50),
    aes(label = contract_number),
    size = 3,
    nudge_y = 0.5,
    show.legend = FALSE
  ) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    color = "grey60"
  ) +
  geom_jitter(
    width = 0.15,
    height = 0.15,
    size = 4,
    alpha = 0.6
  ) +
  scale_x_continuous(
    breaks = scales::breaks_pretty(n = 15)
  ) +
  scale_y_continuous(
    breaks = scales::breaks_pretty(n = 10)
  ) +
  scale_color_manual(
    values = program_colors_high_contrast,
    labels = \(x) tools::toTitleCase(x),
    name = "GCS Item"
  ) +
  labs(
    title = "Modification Counts: Portfolio vs. USASpending",
    x = "Portfolio Modifications",
    y = "USASpending Modifications"
  ) +
  theme_army()


gcs_plots$comparison_summary
```

## Spending trends

-   Award amounts in this data set max at FY2018
-   Abrams by far spent the most
-   Seems Stryker data is limited or there was a pause in acquisition during this period

```{r}
spending <- gcs_portfolio %>%
  select(
    fiscal_year = 5,
    award_amount = 7,
    gcs_item = 15
  ) %>%
  mutate(gcs_item = str_to_title(gcs_item)) %>%
  summarise(
    award_amount = sum(replace_na(award_amount, 0)),
    .by = c(fiscal_year, gcs_item)
  ) %>%
  complete(
    fiscal_year,
    gcs_item,
    fill = list(award_amount = 0)
  ) %>%
  {
    bind_rows(
      .,
      summarise(
        .,
        award_amount = sum(award_amount),
        .by = fiscal_year
      ) %>%
        mutate(gcs_item = "Total")
    )
  } %>%
  mutate(
    label_y = case_when(
      gcs_item == "Total"    ~ award_amount + 20e6,
      gcs_item == "Abrams"   ~ award_amount + 8e6,
      gcs_item == "Bradley"  ~ award_amount + 44e6,
      gcs_item == "Stryker"  ~ award_amount - 10e6,
      TRUE                   ~ award_amount
    )
  )

gcs_plots$spending_trends <-
  ggplot(
    spending,
    aes(
      fiscal_year,
      award_amount,
      color = gcs_item,
      group = gcs_item
    )
  ) +
  geom_line(
    aes(linetype = gcs_item),
    linewidth = 1.2
  ) +
  geom_point(size = 2.5) +
  geom_label(
    aes(
      y = label_y,
      label = scales::label_dollar(
        scale_cut = scales::cut_short_scale()
      )(award_amount),
      fill = gcs_item
    ),
    size = 2.8,
    label.size = 0.15,
    color = "#F2F2F2",   # or "black" if colors are light
    alpha = 1,
    show.legend = FALSE
  ) +
  scale_color_manual(values = program_colors_upper) +
  scale_fill_manual(values = program_colors_upper) +
  scale_linetype_manual(
    values = c(
      Abrams = "solid",
      Bradley = "solid",
      Stryker = "solid",
      Total = "longdash"
    )
  ) +
  scale_y_continuous(
    labels = scales::label_dollar(scale_cut = scales::cut_short_scale()),
    expand = expansion(mult = c(0.05, 0.25))
  ) +
  scale_x_continuous(
    breaks = sort(unique(spending$fiscal_year))
  ) +
  labs(
    title = "GCS Annual Contract Spending",
    subtitle = "FY16-FY20 Award Obligations by Program",
    x = "Fiscal Year",
    y = "Award Amount",
    color = "Program",
    linetype = "Program"
  ) +
  theme_army()

gcs_plots$spending_trends
```

## Key technologies

-   Product and Service Code (PSC) is a four-digit code used by all federal government contracting activities for identifying and classifying the services and Supplies & Equipment (S&E) that are purchased under contract
-   Top 10 PSC by GCS program shows two top PSC shared by the Abrams and Bradley programs
-   More granular detail can be discern if additional time were utilized to scrape PSC definitions and identify key technologies or PSC of interest
-   It is notable the top two PSC codes are shared by the Abrams and Bradley programs

```{r}
# select and count PSCs by GCS program
technologies <- gcs_portfolio %>% 
  select(1, "fiscal_year" = 5, "psc" = 14, 15) %>% 
  mutate(psc_cleaned = str_extract(psc, "\\(([A-Za-z0-9]{4})\\)")) %>%
  filter(!is.na(psc_cleaned)) %>%
  count(gcs_item, psc_cleaned, sort = TRUE) %>%
  group_by(gcs_item) %>%
  slice_max(order_by = n, n = 10, with_ties = FALSE) %>%
  ungroup()

# totals by program
psc_totals <- technologies %>%
  summarise(total = sum(n), .by = gcs_item)

# add totals for facet labels
technologies <- technologies %>%
  left_join(psc_totals, by = "gcs_item")

gcs_plots$psc_code <-
  ggplot(
    technologies,
    aes(x = reorder(psc_cleaned, n), y = n, fill = gcs_item)
  ) +
  geom_col(show.legend = FALSE) +
  facet_wrap(
    ~gcs_item,
    scales = "free_y",
    labeller = as_labeller(\(x) {
      totals <- setNames(psc_totals$total, psc_totals$gcs_item)
      paste0(str_to_title(x), "\n(n = ", scales::comma(totals[x]), ")")
    })
  ) +
  coord_flip() +
  scale_fill_manual(values = program_colors) +
  labs(
    x = "Product & Service Code (PSC)",
    y = "Contract Count",
    title = "Top 10 PSC by GCS Program"
  ) +
  geom_text(
    aes(label = scales::comma(n)),
    hjust = -0.15,
    size = 3
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_army()


gcs_plots$psc_code
```

## Contract share

### All Programs

```{r}
# create agency share table
program_summary <- gcs_portfolio %>%
  select(
    gcs_item = 15,
    contracting_agency = 9
  ) %>%
  filter(!is.na(contracting_agency)) %>%
  group_by(gcs_item) %>%
  summarise(
    total_contracts = n(),
    unique_agencies = n_distinct(contracting_agency),
    primary_agency = names(sort(table(contracting_agency), decreasing = TRUE))[1],
    primary_agency_share = max(table(contracting_agency)) / n(),
    .groups = "drop"
  ) %>% 
  mutate(gcs_item = str_to_title(gcs_item))

gcs_plots$agency_share <- ggplot(
  program_summary,
  aes(
    x = reorder(gcs_item, primary_agency_share),
    y = primary_agency_share,
    fill = gcs_item
  )
) +
  geom_col(show.legend = FALSE) +
  geom_text(
    aes(label = scales::percent(primary_agency_share, accuracy = 1)),
    hjust = -0.1,
    size = 5
  ) +
  scale_fill_manual(values = program_colors_upper) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1.1)
  ) +
  coord_flip() +
  labs(
    title = "GCS Programs Contracting Concentration",
    subtitle = "Distribution of contract actions by largest contracting agency (DoD)",
    x = NULL,
    y = "Share of Contract Actions"
  ) +
  theme_army()

gcs_plots$agency_share
```

### Abrams

```{r}
abrams_agency <- gcs_portfolio %>%
  filter(str_to_title(gcs_item) == "Abrams") %>%
  select(
    contracting_agency = 9
  ) %>%
  filter(!is.na(contracting_agency)) %>%
  count(contracting_agency) %>%
  mutate(
    share = n / sum(n),
    contracting_agency = forcats::fct_reorder(contracting_agency, share)
  )

gcs_plots$abrams_agency_share <-
  ggplot(
    abrams_agency,
    aes(
      x = share,
      y = contracting_agency
    )
  ) +
  geom_col(
    fill = "#4B5320",
    width = 0.7
  ) +
  geom_text(
    aes(
      label = scales::percent(share, accuracy = 0.1)
    ),
    hjust = -0.15,
    size = 4
  ) +
  scale_x_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.1))
  ) +
  labs(
    title = "Abrams Contracting Concentration",
    subtitle = "Distribution of contract actions by awarding agency",
    x = "Share of Contract Actions",
    y = NULL
  ) +
  theme_army()

gcs_plots$abrams_agency_share
```

## Vendors

-   General Dynamics Land Systems Inc. (GDLS) was by far the highest awarded vendor in this time period
-   Abrams awards dominated for the top 10 vendors by total award amount in this period
-   It is interesting that the majority of vendors only contracted for one of the two programs

```{r}
vendor_summary <- gcs_portfolio %>% 
  select(
    award_amount = 7,
    vendor = 12,
    gcs_item = 15
  ) %>%
  mutate(
    award_amount = replace_na(award_amount, 0)
  ) %>%
  group_by(vendor, gcs_item) %>%
  summarise(
    total_awards = sum(award_amount),
    .groups = "drop"
  ) %>%
  group_by(vendor) %>%
  mutate(
    vendor_total = sum(total_awards)
  ) %>%
  ungroup() %>%
  slice_max(vendor_total, n = 10) 

# calculate vendor totals for labels
vendor_labels <- vendor_summary %>%
  group_by(vendor) %>%
  summarise(
    vendor_total = sum(total_awards),
    .groups = "drop"
  )

gcs_plots$vendors <- ggplot(
  vendor_summary,
  aes(
    x = reorder(vendor, vendor_total),
    y = total_awards,
    fill = gcs_item
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    data = vendor_labels,
    aes(
      x = vendor,
      y = vendor_total,
      label = scales::label_dollar(
        scale_cut = scales::cut_short_scale()
      )(vendor_total)
    ),
    inherit.aes = FALSE,
    hjust = -0.1,
    size = 4
  ) +
  coord_flip() +
  scale_fill_manual(
    values = program_colors_high_contrast,
    labels = c(
      abrams = "Abrams",
      bradley = "Bradley",
      stryker = "Stryker"
    )
  ) +
  scale_y_continuous(
    labels = scales::label_dollar(
      scale_cut = scales::cut_short_scale()
    ),
    expand = expansion(mult = c(0, 0.2))
  ) +
  labs(
    title = "Top Vendors by GCS Award Amount (FY16-FY20)",
    x = NULL,
    y = "Award Amount",
    fill = "Program"
  ) +
  theme_army()

gcs_plots$vendors
```

# Print

```{r}
fig_path <- "figures"

#dir.create(fig_path, showWarnings = FALSE)

purrr::iwalk(
  gcs_plots,
  ~ ggsave(
      filename = file.path(fig_path, paste0(.y, ".png")),
      plot = .x,
      width = 8,
      height = 6,
      dpi = 300
    )
)
```
